\Title

### 2022-09-16 Grimaldi-etal-BiolCybernetics.tex
new figures

\newcommand{\Acknowledgments}{This research was funded by the European Union ERA-NET CHIST-ERA 2018 research and innovation program under grant agreement N° ANR-19-CHR3-0008-03 (``\href{http://aprovis3d.eu/}{APROVIS3D}''). %HL and LP received funding from the ANR project  N° ANR-20-CE23-0021 (''\href{https://laurentperrinet.github.io/grant/anr-anr/}

@@

%%%-----------------------------------------------------------------
\begin{figure}[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion-task.pdf}
    \note{TODO: redo figure according to caption}
    \caption{
    {\bf Motion detection task.} {\bf (a)} To mimic the effect of a saccadic-like movement, the axis of view moves following a step-like random walk, and we show here on example path of the trajectory. {\bf (b)} We use large-scale natural images ($512\times512$) in which an aperture ($128\times128$) extracts a sub-image image in the axis of view (here the cross at the center of the image) such as to reproduce the effect of displacing the eye. {\bf (c)} Recording the dynamics of the sub-image as a function of time, it generates a naturalistic movie which may be transformed to an event-based representation. Mimicking the retina, this representation codes for proportional increments or decrements of the luminance in the image, respectively ON (in red) and OFF (in blue) events. This will constitute the input to the spiking neural network.}
    \label{fig:motion-task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye movement trajectories are generated, we can apply them to a visual scene. For this purpose, we selected a database of 10 natural grayscale images that are commonly used to study the statistics of natural images~\citep{olshausen_emergence_1996}. Note that these are pre-processed to equalize the energy in each frequency band (i.e., whitened). This process is known to occur as early as the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $512 \times 512$ in size and we will extract sub-images of size $128 \times 128$ which will be positioned around the center of the gaze at each time step (see \seeFig{motion-task}-(b)). We will discretize the time in $1~\ms$ bins and produce movies of duration N-T = $250~\ms$. To avoid border effects, we will position a draw of the complete trajectory at random in the image space so that the sub-image is translated using the position given by the trajectory at each time step. The translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the velocity such that a velocity of $V=1$ corresponds to a movement of one pixel per frame (i.e., per time bin).

@@

    \includegraphics[width=\linewidth]{figures/3D_kernels_oneline.png}

@@

Let us highlight some innovations in the contributions presented in this paper. First, the generic heterosynaptic model is formalized from first principles for optimal detection of the event-based spatiotemporal motifs, whereas~\citet{ghosh_spatiotemporal_2019} use a correlation-based heuristic, which we have proven to be less efficient. Moreover, in comparison to HOTS~\citep{lagorce_hots_2017} the weights are explainable as they directly inform on the logit (inverse sigmoid of the probability) of detecting each spatio-temporal spiking motif. Another novelty is that the model simultaneously learns the weights and the delays, while for example the polychronization model~\citep{izhikevich_polychronization_2006} only learns the weights using STDP while the delays are randomly drawn and their values frozen. Moreover, the model is evaluated on a realistic task, while models such as tempotron are tested on simplified problems~\citep{gutig_tempotron_2006}. The main contribution is to provide a model that is suitable for any type of spatio-temporal spiking motifs and that can be trained in a supervised way by providing a dataset of supervision pairs. This allows for a more flexible definition of the model using this properly labeled dataset instead of relying on a careful description of the physical rules governing a task, e.g., the luminance conservation principle for motion detection~\citep{benosman_asynchronous_2012, dardelet_event-by-event_2021}.
%
\subsection{Main limits}
%%%-----------------------------------------------------------------
We have identified several limits to our model, the most important being now detailed. First, the complete framework is based on a discrete binning of time that is incompatible with the continuous nature of biological time. We have used this binning to be able to efficiently implement the framework on conventional hardware, especially GPUs, and in particular to be able to use three-dimensional convolutions. We have tested the effect of the size of the time bin and shown that it has essentially no impact on the results presented in this paper. This is consistent with the relative robustness of other event-based frameworks such as HOTS~\citep{lagorce_hots_2017}, where the accuracy was not affected when the input spikes were subjected to noisy perturbations up to $1~\ms$~\citep{grimaldi_robust_2022}. This suggests the possibility of analytically including a precision term in the temporal value of the input spikes, a mechanism potentially implemented by the filtering that is implemented by the synaptic time constant of about $5~\ms$. Furthermore, it is possible to circonvent the necessity of using a time discretization by using a purely event-based scheme. Indeed, it is not necessary to compute voltage traces between two spikes~\citep{hanuschkin_general_2010} and it is thus possible to define a purely event-based framework. Such an architecture could provide promising speed gains for the calculations.

@@

The model is trained on a low-level local motion detection task, and one might wonder if it could be trained on higher-level tasks. An example task would be depth estimation in the visual scene. There are multiple sources of information for inferring depth, such as binocular disparity or changes in texture or shading, but in our case, motion parallax would be the most significant cue~\citep{rogers_motion_1979}. This is because objects close to an observer move relatively faster on the retina than an object at a great distance, and also that visual occlusions depend on depth ordering. Using this information, one can segment objects and estimate their depth~\citep{yoonessi_contribution_2011}. However, this would require computing first the optical flow. A possible implementation is therefore to add a new layer to our model, in analogy with the hierarchical organization highlighted in the visual cortex. This is theoretically possible by using the output of our model (which estimates velocity in retinotopic space) as input to a new layer of neurons that would estimate velocity in the visual field, including the depth dimension in the output supervision labels. This could have direct and important applications, for example in autonomous driving, to detect obstacles in a fast and robust way. Another extension would be to actively generate sensor movements (physically or virtually) to yield better depths estimates, notably to disambiguate uncertain estimates~\citep{nawrot_eye_2003}.

>>>>

\newcommand{\Acknowledgments}{
Thanks to CB, TS, ...
This research was funded by the European Union ERA-NET CHIST-ERA 2018 research and innovation program under grant agreement N° ANR-19-CHR3-0008-03 (``\href{http://aprovis3d.eu/}{APROVIS3D}''). %HL and LP received funding from the ANR project  N° ANR-20-CE23-0021 (''\href{https://laurentperrinet.github.io/grant/anr-anr/}

@@

%%%-----------------------------------------------------------------
\begin{figure}[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \note{TODO: redo figure according to caption}
    \caption{
    {\bf Motion detection task.} {\bf (a)} To mimic the effect of a saccadic-like movement, the axis of view moves following a step-like random walk, and we show here on example path of the trajectory. {\bf (b)} We use large-scale natural images ($512\times512$) in which an aperture ($128\times128$) extracts a sub-image image in the axis of view (here the cross at the center of the image) such as to reproduce the effect of displacing the eye. {\bf (c)} Recording the dynamics of the sub-image as a function of time, it generates a naturalistic movie which may be transformed to an event-based representation. Mimicking the retina, this representation codes for proportional increments or decrements of the luminance in the image, respectively ON (in red) and OFF (in blue) events. This will constitute the input to the spiking neural network.}
    \label{fig:motion-task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye movement trajectories are generated, we can apply them to a visual scene. For this purpose, we selected a database of 10 natural grayscale images that are commonly used to study the statistics of natural images~\citep{olshausen_emergence_1996}. Note that these are pre-processed to equalize the energy in each frequency band (i.e., whitened). This process is known to occur as early as the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $512 \times 512$ in size and we will extract sub-images of size $128 \times 128$ which will be positioned around the center of the gaze at each time step (see \seeFig{motion-task}-(b)). We will discretize the time in $1~\ms$ bins and produce movies of duration N-T = $250~\ms$. To avoid border effects, we will position a draw of the complete trajectory at random in the image space so that the sub-image is translated using the position given by the trajectory at each time step. The translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the velocity such that a velocity of $V=1$ corresponds to a movement of one pixel per frame (i.e., per time bin).

@@

    \includegraphics[width=\linewidth]{figures/motion_kernels.pdf}

@@

Let us highlight some innovations in the contributions presented in this paper. First, the generic heterosynaptic model is formalized from first principles for optimal detection of the event-based spatiotemporal motifs, whereas~\citet{ghosh_spatiotemporal_2019} use a correlation-based heuristic, which we have proven to be less efficient. Moreover, in comparison to HOTS~\citep{lagorce_hots_2017} the weights are explainable as they directly inform on the logit (inverse sigmoid of the probability) of detecting each spatio-temporal spiking motif. Another novelty is that the model simultaneously learns the weights and the delays, while for example the polychronization model~\citep{izhikevich_polychronization_2006} only learns the weights using STDP while the delays are randomly drawn and their values frozen. Moreover, the model is evaluated on a realistic task, while models such as tempotron are tested on simplified problems~\citep{gutig_tempotron_2006}. The main contribution is to provide a model that is suitable for any type of spatio-temporal spiking motifs and that can be trained in a supervised way by providing a dataset of supervision pairs. This allows for a more flexible definition of the model using this properly labeled dataset instead of relying on a careful description of the physical rules governing a task, e.g., the luminance conservation principle for motion detection~\citep{benosman_asynchronous_2012, dardelet_event-by-event_2021}.
%
\subsection{Main limits}
%%%-----------------------------------------------------------------
We have identified several limits to our model, the most important being now detailed. First, the complete framework is based on a discrete binning of time that is incompatible with the continuous nature of biological time. We have used this binning to be able to efficiently implement the framework on conventional hardware, especially GPUs, and in particular to be able to use three-dimensional convolutions. We have tested the effect of the size of the time bin and shown that it has essentially no impact on the results presented in this paper. This is consistent with the relative robustness of other event-based frameworks such as HOTS~\citep{lagorce_hots_2017}, where the accuracy was not affected when the input spikes were subjected to noisy perturbations up to $1~\ms$~\citep{grimaldi_robust_2022}. This suggests the possibility of analytically including a precision term in the temporal value of the input spikes, a mechanism potentially implemented by the filtering that is implemented by the synaptic time constant of about $5~\ms$. Furthermore, it is possible to circonvent the necessity of using a time discretization by using a purely event-based scheme. Indeed, it is not necessary to compute voltage traces between two spikes~\citep{hanuschkin_general_2010} and it is thus possible to define a purely event-based framework. Such an architecture could provide promising speed gains for the calculations.  \note{pourrait parler du poster bernstein? de nadafian?}

@@

The model is trained on a low-level local motion detection task, and one might wonder if it could be trained on higher-level tasks. An example task would be depth estimation in the visual scene. There are multiple sources of information for inferring depth, such as binocular disparity or changes in texture or shading, but in our case, motion parallax would be the most significant cue~\citep{rogers_motion_1979}. This is because objects close to an observer move relatively faster on the retina than an object at a great distance, and also that visual occlusions depend on depth ordering. Using this information, one can segment objects and estimate their depth~\citep{yoonessi_contribution_2011}. However, this would require computing first the optical flow, that is extending the framework described here for a rigid full-field motion to a more generic one where motion may vary in the visual field. A possible implementation is therefore to add a new layer to our model, in analogy with the hierarchical organization highlighted in the visual cortex. This is theoretically possible by using the output of our model (which estimates velocity in retinotopic space) as input to a new layer of neurons that would estimate velocity in the visual field, including the depth dimension in the output supervision labels. This could have direct and important applications, for example in autonomous driving, to detect obstacles in a fast and robust way. Another extension would be to actively generate sensor movements (physically or virtually) to yield better depths estimates, notably to disambiguate uncertain estimates~\citep{nawrot_eye_2003}.
